---
title: "p8105 Homework 2"
output: github_document
---

```{r}
library(tidyverse)
options(tibble.print_min = 5)
```

# Problem 1

Read and clean the data; retain line, station name, station latitude / longitude, routes served, entry, vending, entrance type, and ADA compliance. Convert the entry variable from character (YES vs NO) to a logical variable.

```{r}
nyc_transit = 
  read_csv("./data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv",
           col_types = cols(Route8 = "c", Route9 = "c", Route10 = "c", Route11 = "c")) %>% 
  janitor::clean_names() %>% 
  select(
    line, starts_with("station"), starts_with("route"), entry, exit_only, vending,          entrance_type, ada) %>% 
  mutate(entry = ifelse(entry == "YES", TRUE, FALSE))

nyc_transit
```

The raw dataset constains 32 columns, including line, station name, station latitude / longitude, routes served, entry, vending, entrance type, corner, etc,. There are `r nrow(nyc_transit)` rows * `r ncol(nyc_transit)` columns in the resulting dataset.

The process begins with data import, updates variable names, and selects the columns that will be used in later parts fo this problem. We update `entry` from `yes` / `no` to a logical variable. As part of data import, we specify that `Route` columns 8-11 should be character for consistency with 1-7. 

As it stands, these data are not "tidy": route number should be a variable, as should route. That is, to obtain a tidy dataset we would need to convert `route` variables from wide to long format. This will be useful when focusing on specific routes, but may not be necessary when considering questions that focus on station-level variables.

Then we are going to answer these questions:

* How many distinct stations are there?

The following code chunk selects station name and line, and then uses `distinct()` to obtain all unique combinations. As a result, the number of rows in this dataset is the number of unique stations.

```{r}
nyc_transit %>% 
  select(line, station_name) %>% 
  distinct
```

There are 465 distinct stations.

* How many stations are ADA compliant?

This produces a dataframe in which the number of rows is the number of ADA compliant stations. 

```{r}
nyc_transit %>% 
  filter(ada == TRUE) %>% 
  select(line, station_name) %>% 
  distinct
```

84 stations are ADA compliant.

* What proportion of station entrances / exits without vending allow entrance?

To compute the proportion of station entrances / exits without vending allow entrance, we first exclude station entrances that do not allow vending. Then, we focus on the `entry` variable -- this logical, so taking the mean will produce the desired proportion.

```{r}
nyc_transit %>% 
  filter(vending == "NO") %>% 
  pull(entry) %>% 
  mean
```

37.7% of station entrances / exits without vending allow entrance.

Reformat data so that route number and route name are distinct variables. How many distinct stations serve the A train? Of the stations that serve the A train, how many are ADA compliant?

We convert the dataset from wide to long format, and then use `filter`, `select`, and `distinct` statements to obtain dataframes with the required stations in rows.

```{r}
nyc_transit %>% 
  pivot_longer(
    route1:route11,
    names_to = "route_num",
    values_to = "route") %>% 
  filter(route == "A") %>% 
  select(station_name, line) %>% 
  distinct
```

There are 60 distinct stations serving the A train.

```{r}
nyc_transit %>% 
  pivot_longer(
    route1:route11,
    names_to = "route_num",
    values_to = "route") %>% 
  filter(route == "A" & ada == TRUE) %>% 
  select(station_name, line) %>% 
  distinct
```

Of the stations that serve the A train, 17 stations are ADA compliant.


# Problem 2

Firstly we read and clean the Mr. Trash Wheel sheet:

```{r}
mr_tw = 
  readxl::read_excel("./data/Trash Wheel Collection Data.xlsx", 
                     sheet = "Mr. Trash Wheel", range = "A2:N549") %>%
  janitor::clean_names() %>% 
  filter(!is.na(dumpster)) %>% 
  mutate(
    sports_balls = as.integer(round(sports_balls, 0)),
    trash_wheel = "Mr.",
    year = as.integer(year)
    )

mr_tw
```

Then use a similar process to import, clean, and organize the data for Professor Trash Wheel:

```{r}
prof_tw = 
  readxl::read_excel("./data/Trash Wheel Collection Data.xlsx", 
                     sheet = "Professor Trash Wheel", range = "A2:M96") %>%
  janitor::clean_names() %>% 
  filter(!is.na(dumpster)) %>% 
  mutate(
    trash_wheel = "Professor",
    year = as.integer(year)
    )

prof_tw
```

Combine those two dataframs to produce a single tidy dataset.

```{r}
tw_tidy = 
  bind_rows(mr_tw, prof_tw) %>% 
  select(trash_wheel, everything())

tw_tidy
```

The tw_tidy dataset is the combination of **Mr. Trash Wheel** and **Professor Trash Wheel**. There are 641 observations in total. The variables "trash_wheel" (__Mr.__ and __Professor__) make each observation distinct. And there are 15 columns, of which the "sports_balls" variable only works for Mr. Trash Wheel. Besides, other columns include year, date, glass bottles, grocery bags, home powered, etc,.

The following chunks of code are going to answer these questions: for available data, what was the total weight of trash collected by Professor Trash Wheel? What was the total number of sports balls collected by Mr. Trash Wheel in 2020?

```{r}
tw_tidy %>% 
  filter(trash_wheel == "Professor") %>% 
  select(weight_tons) %>% 
  sum
```

The total weight of trash collected by Professor Trash Wheel is 190.12 tons.

```{r}
tw_tidy %>% 
  filter(trash_wheel == "Mr." & year == 2020) %>% 
  select(sports_balls) %>% 
  sum
```

The total number of sports balls collected by Mr. Trash Wheel in 2020 is 856.


# Problem 3

First, clean the data in pols-month.csv. Use `separate()` to break up the variable `mon` into integer variables `year`, `month`, and `day`; replace month number with month name; create a `president` variable taking values `gop` and `dem`, and remove `prez_dem` and `prez_gop`; and remove the day variable.

```{r}
pols_df = read_csv("./data/fivethirtyeight_datasets/pols-month.csv") %>% 
  janitor::clean_names() %>% 
  separate(mon, into = c("year", "month", "day"), sep = "-") %>% 
  mutate(
    month = month.name[as.integer(month)],
    year = as.integer(year),
    president = ifelse(prez_gop == 1, "gop", "dem")
  ) %>% 
  select(-prez_dem, -prez_gop, -day)

pols_df
```

Second, clean the data in snp.csv using a similar process to the above. For consistency across datasets, arrange according to year and month, and organize so that `year` and `month` are the leading columns.

```{r}
snp_df = read_csv("./data/fivethirtyeight_datasets/snp.csv") %>% 
  janitor::clean_names() %>%
  mutate(
    date = lubridate::parse_date_time2(date, orders = "mdy", cutoff_2000 = 23)
  ) %>% 
    separate(date, into = c("year", "month", "day"), sep = "-") %>% 
  mutate(
    month = as.integer(month),
    year = as.integer(year)
  ) %>% 
  arrange(year, month) %>% 
  mutate(month = month.name[month]) %>% 
  select(year, month, close)

snp_df
```

Third, tidy the unemployment data so that it can be merged with the previous datasets. This process will involve switching from “wide” to “long” format; ensuring that key variables have the same name; and ensuring that key variables take the same values.

```{r}
unemployment = read_csv("./data/fivethirtyeight_datasets/unemployment.csv") %>%
  rename(year = Year) %>% 
  pivot_longer( 
    Jan:Dec,
    names_to = "month", 
    values_to = "unemployment") %>% 
  mutate(
    month = plyr::mapvalues(month, from = month.abb, to = month.name),
    year = as.integer(year))

unemployment
```

Join the datasets by merging `snp` into `pols`, and merging `unemployment` into the result.

```{r}
result1 = full_join(pols_df, snp_df)
final_version = full_join(result1, unemployment)
final_version
```

The first raw dataset pols-month.csv contains 9 variablea. The `mon` indicates the year, month, and day. Other columns such as gov_gop and sen_dem are included. We created a `president` variable taking values gop and dem. The second dataset `snp` also has the `date` variable with 2-digit years. It has another column named `close`. The `unemployment` dataset has the `Year` variable and other 12 columns named by 12 months. This table reports the unemployment rate of each month in every year. We converted this dataset from `wide` to `long`.

We got the resulting dataset `final_version`. It has `r nrow(final_version)` rows * `r ncol(final_version)` columns. The year range is from 1947 to 2015. There are some missing values at the last few months in 2015. In addition, there is no `close` data from 1947 to 1949, and no `unemployment` data in 1947. The resulting dataset is merged from the 3 datasets above by `year` and `month`. Variables `gov_gop`, `sen_gop`, `rep_gop`, `gov_dem`, `sen_dem`, `rep_dem`, `president`, `close`, and `unemployment` are retained.
